---
title: "COVIDPostiveTestingSA"
author: "Belay Birlie Yimer, ..., Ziv Shkedy"
date: "02/03/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

- Individuals rely on test results to guide their medical treatment and decisions on whether to self-isolate.
- Public health officials rely on the results to track the state of the pandemic, and policymakers use this information to guide decisions on reopening schools and businesses.
- One number—the “percent positive”—is often cited in these decisions.
- The percent positive is the percentage of all coronavirus tests performed that are actually positive, or: (positive tests)/(total tests) x 100%.

- The percent positive helps public health officials answer questions such as:
    - What is the current level of SARS-CoV-2 (coronavirus) transmission in the community?
    - Are we doing enough testing for the amount of people who are getting infected?


# Methods

## Data
We downloaded data from Coronavirus COVID-19 (2019-nCoV) Data Repository for South Africa maintained by Data Science for Social Impact research group at the University of Pretoria [ref]. The data repository captures the daily number of new cases, number of tests, number of deaths and recoveries. Our primary outcome of interest was the daily number of newly diagnosed COVID-19 cases and the unit of time used in modelling was a day. We used the daily case reports from March 12, 2020, until February 27, 2021, in our analysis.

## Statistical analysis 
We assume the number of postive cases to have a binomial distribution. Let $\pi_t$ denote the daily postive testing rate per test, $Y_t$ be the daily number of confirmed COVID-19 cases out of the $n_t$ indiviudals tested at day $t$. Our aim is to model the probability $\pi_t$ over time and produce a model-based estimate for it's first-order derivative. The model for the daily poitive testing rate is then given by 
\[
\begin{aligned}
Y_t \sim& Binomial(n_t, \pi_t),\ \ t=1,...,T,\\
logit(\pi_t)=&f(t).
\end{aligned}
\]

The smooth function $f(t)$ is represtend by basis expansion of rank $K$. That is, 
\[
f(t)=\sum_{i=1}^{K}\beta_ib_i(t)
\]
where $b_i(t)$s are set of thin plate spline basis functions and $\beta_i$s are unknown coefficients of the basis functions. The spline model parameters can be estimated by considering the goodness-of-fit and the degree of smoothness. Writing all the basis coefficents in one parameter vector $\boldsymbol{\beta}$, 

\[
\begin{aligned}
l(\beta)-\frac{\lambda}{2} \sum_{j=1}^M\beta^TS\beta\
\end{aligned}
\]
where $S$ is metrices of known coefficents and $\lambda$ is the smoothing parameter to be estimated. 

# Results 

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message = FALSE, warning = FALSE}
library(ggplot2)
library(visreg)
library(cowplot)
#'library(visibly)
library(tidyverse)
library(gridExtra)
library(ggpubr)
library(patchwork)
library(lubridate)
library(kableExtra)
library(data.table)
library(sfsmisc)
library(brms)
```

```{r message = FALSE, warning = FALSE}
#'------------- Load data 
tests_sa <- read.csv("https://raw.githubusercontent.com/dsfsi/covid19za/master/data/covid19za_timeline_testing.csv",header=TRUE)
cases_sa <- read.csv("https://raw.githubusercontent.com/dsfsi/covid19za/master/data/covid19za_provincial_cumulative_timeline_confirmed.csv",header=TRUE)
#'------------- Prepare the two data sets to merge  
tests_sa<-tests_sa%>%mutate(date2=lubridate::ymd(YYYYMMDD))
cases_sa<-cases_sa%>%mutate(date2=lubridate::ymd(YYYYMMDD))

variable_keep_test<-c("date2","cumulative_tests")
variable_keep_cases<-c("date2","total")

sa_cov_dat<-left_join(cases_sa[,variable_keep_cases], tests_sa[,variable_keep_test], by="date2")
# sa_cov_dat[18:35,] also unknown number of tests on 25-03, and 07-04 
sa_cov_dat<-sa_cov_dat[sa_cov_dat$date2>'2020-03-09'&!is.na(sa_cov_dat$cumulative_tests),]
sa_cov_dat<-sa_cov_dat[sa_cov_dat$date2<'2021-02-28',]#temp


# You need to start from 2020-03-12 in order to properly calaculate daily cases and tests
# Compute daily cases and tests 
sa_cov_dat<-sa_cov_dat%>%mutate(Dialy_cases=-1*(lag(total)-total), Daily_tests=-1*(lag(cumulative_tests)-cumulative_tests))
sa_cov_dat<-sa_cov_dat[-1, ]


```
Figure 1 presents the daily number of reported COVID-19 cases from 12 March 2020 to 27 February 2021.  Similar to elsewhere in the world,  South Africa pass through a two-wave pandemic. The pandemic's first peak was on 07 July 2020, where up to 13944 new COVID-19 cases reported, followed by a second peak in January 2021, where more than 21,000 daily cases reported. 
Figure 2 presents the cumulative number of new reported COVID-19 cases and tests performed. To date, 8,838,937 tests have been conducted, and a total of 1,500,677 cases reported.



```{r daily-cases,out.width = "99%",fig.cap = "Daily number of COVID-19 cases in South Africa from 12/03/2020-27/02/2021."}

p1<-ggplot(sa_cov_dat, aes(x=date2, y=Dialy_cases,group=1))+geom_point()+xlab("Date")+ylab("Daily cases")+theme_bw()
p1

```

```{r cummulative,out.width = "99%",fig.cap = "The cummulative number of COVID-19 cases and Cummulative number of tests in South Africa from 12/03/2020-27/02/2021. Red-line denote the number of cases and blue-line denotes the number of tests."}

dataa<-gather(sa_cov_dat[,c("date2", "total", "cumulative_tests")], Outcome, count, total:cumulative_tests, factor_key=TRUE)
dataa<-dataa %>% 
  mutate(Type = case_when(
    Outcome=="cumulative_tests" ~ 'Tests',
    TRUE ~ 'Cases') )

p2<-ggplot(dataa,aes(x=date2, y=count,group=Type))+
  geom_line(aes(color=Type))+
  xlab("Date")+ylab("Cumulative count")+
theme_bw()+theme(legend.position="none")

p2
#p2<-ggplot(sa_cov_dat, aes(x=date2, y=total,group=1))+geom_line(size = 1.5)+xlab("Date")+ylab("Cummulative cases")+theme_bw()


#p3<-ggplot(sa_cov_dat, aes(x=date2, y=cumulative_tests,group=1))+geom_line(size = 1.5)+xlab("Date")+ylab("Cummulative tests")+theme_bw()

#p1_3<-p1/p2

#p1_3


```


```{r message = FALSE, warning = FALSE}
sa_cov_dat<-sa_cov_dat%>%mutate(time=cumsum(c(0,diff.Date(date2))))

#fit1<- brm(bf(Dialy_cases|trials(Daily_tests) ~ s(time)),
#                data = sa_cov_dat, family = binomial(), cores = 2, seed = 17,
#                iter = 6000, warmup = 3000, thin = 10, refresh = 0,
#                control = list(adapt_delta = 0.99))
#saveRDS(fit1, file = "fit1.rds")
fit1<-readRDS("fit1.rds")

```


```{r ,fig.height= 8,dpi = 300,fig.cap = "Upper Panel: Observed postive testing rate. Middle panel: The linear predictor of the smoother with 95% creadible interval. Lower panel: The derivative of the estimated probability of postive testing with the 95% creadible interval."}

#----------------- get the profile plot - liner predictor 
len = 345
x1_new <- seq(min(sa_cov_dat$time), max(sa_cov_dat$time), length.out = len)
newdata<-data.frame(x1_new, rep(1, 345))
names(newdata)<-c('time','Daily_tests')

predictions <- brms::posterior_linpred(fit1, newdata = newdata, summary = FALSE)
means <- apply(predictions, MARGIN = 2,mean)
lower <- apply(predictions, MARGIN = 2,quantile, prob = 0.055)
upper <- apply(predictions, MARGIN = 2,quantile, prob = 0.945)
Date2<-sa_cov_dat$date2
dim(sa_cov_dat)
profile_plot<-data.frame(means) %>%
  cbind(lower) %>%
  cbind(upper) %>%
  cbind(newdata) %>%
  cbind(Date2)%>%
  ggplot(.,aes(Date2, means)) + geom_line() + 
  geom_ribbon(aes(ymin = lower, ymax = upper), color="blue", alpha = 0.1)+
  ylab("Linear predictor")+xlab("Date")+theme_classic()

# --------------- get the profile plot on the response scale 

#---------------- get the first derivative 
len = 345
epsilon <- 1e-6
x1_new <- seq(min(sa_cov_dat$time), max(sa_cov_dat$time), length.out = len)
first<-data.frame(x1_new, rep(1, len))
names(first)<-c('time','Daily_tests')

second <- first %>%
  mutate(time = time + epsilon)

## get predictions
first_preds <-
  posterior_linpred(fit1,
                    newdata = first,
                    summary = FALSE,)
second_preds <-
  posterior_linpred(fit1,
                    newdata = second,
                    summary = FALSE)


## Calcualte the differennce and divide by epsilon - this is analgous to the dx/dt 
diff <- (second_preds - first_preds) / epsilon


## using the posterior samples, we calculate the mean and lower and upper quantiles
mean_finite_diff  <- apply(diff, MARGIN = 2, FUN = mean)
lower_finite_diff <- apply(diff, MARGIN = 2, FUN = quantile,prob = 0.025)
upper_finite_diff <- apply(diff, MARGIN = 2, FUN = quantile,prob = 0.975)

derivative_plot<-data.frame(mean_finite_diff) %>%
  cbind(lower_finite_diff) %>%
  cbind(upper_finite_diff) %>%
  cbind(first) %>%
  cbind(Date2) %>%
  ggplot(.,aes(Date2, mean_finite_diff)) + geom_line()  +
  geom_ribbon(aes(ymin = lower_finite_diff, ymax = upper_finite_diff), color="blue",alpha = 0.1) + 
  labs(y = "Derivative")+xlab("Date")+
  geom_hline(yintercept = 0, color="red", linetype="dashed") +theme_classic()

postive_rate_plot<-ggplot(sa_cov_dat, aes(x=date2, y= Dialy_cases/Daily_tests,group=1))+geom_point(size = 0.1, stroke = 0, shape = 16)+xlab("Date")+ylab("Postive testing rate")+theme_classic()

pp1<-postive_rate_plot/profile_plot/derivative_plot# work on getting dates right 

ggsave("C:/Users/user/Dropbox (The University of Manchester)/covid_PhD/COVIDPostiveTestingSA/testingplot.png",pp1)
pp1
```